---
title: "R for OHDSI"
author: "Erik Westlund"
date: "2025-08-29"
---

# R for OHDSI

## About Me

- Data Scientist at Johns Hopkins Bloomberg School of Public Health
- Johns Hopkins Biostatistics Center, Department of Biostatistics
- Trained in social sciences with 10+ years experience as data scientist and software developer
- Email: ewestlund@jhu.edu

## Overview

- Why R matters for OHDSI
- Essential R concepts for OHDSI work
- Using AI effectively with R
- Key R skills for debugging and analysis

## About This Presentation

- Focus on practical R skills for OHDSI studies
- Essential knowledge for working with HADES packages
- Tips for debugging and troubleshooting

# Why R in OHDSI?

## OHDSI is Built on R

- **HADES packages** are written in R
- **Strategus** execution framework uses R
- **Study packages** distributed as R code
- Most analytical methods implemented in R

## You Need R To:

- **Debug** when things go wrong
- **Understand** what packages are doing
- **Customize** analyses for your needs
- **Run** network studies at your site

# AI and R

## AI Can Help, But...

- AI tools (ChatGPT, Claude, Copilot) can accelerate coding
- Excellent for boilerplate and common patterns
- Helpful for understanding error messages

## Critical: Test Everything

- **Always test AI-generated code**
- **Examine outputs carefully**
- **Understand what the code does**
- AI doesn't know your specific OMOP CDM structure
- May hallucinate functions that don't exist

## Best Practices with AI

::: {.columns}
::: {.column width="50%"}
### Do's
- Use for learning R concepts
- Debug error messages
- Generate starter code
- Explain existing code
:::

::: {.column width="50%"}
### Don'ts
- Trust blindly
- Skip testing
- Assume it knows OHDSI specifics
- Use without understanding
:::
:::

# Essential R Concepts

## R and RStudio

- **R** is the programming language
- **RStudio** is an IDE (Integrated Development Environment)
- You write R code in RStudio (or VSCode or your preferred editor)
- All OHDSI packages run in this environment

## R Projects

- Always use R Projects for organization
- Sets working directory automatically
- Keeps code and data together
- Essential for reproducibility

```r
# Create new project:
File → New Project
```

## Objects and Assignment

```{r}
#| echo: true
# Assignment uses <- (preferred) or =
cohortId <- 1
targetCohort <- 100
print(cohortId)
print(targetCohort)

# Everything is an object
#
# connectionDetails <- createConnectionDetails(
#   dbms = "postgresql",
#   server = "localhost/ohdsi"
# )
```

## Vectors: R's Building Blocks

```{r}
#| echo: true
# Single values are 1-element vectors
patientCount <- 1000
print(patientCount)

# Multiple values
cohortIds <- c(1, 2, 3, 4, 5)
conceptIds <- c(313217, 313236, 313238)
print(cohortIds)
print(conceptIds)
```

## Named Vectors

```{r}
#| echo: true
# Named vectors are very useful in OHDSI
cohortNames <- c(
  "diabetes" = 1,
  "hypertension" = 2,
  "depression" = 3
)
print(cohortNames)

# Access by name
print(cohortNames["diabetes"])
```

## Data Frames

- Most OHDSI results are data frames
- Tables with rows and columns
- Each column can be different type

## Data Frame Output

```{r}
#| echo: true
# Example cohort table
library(kableExtra)
cohort <- data.frame(
  cohortDefinitionId = c(1, 1, 2, 2),
  subjectId = c(100, 101, 100, 102),
  cohortStartDate = as.Date(c("2020-01-01", "2020-02-01", 
                              "2020-03-01", "2020-04-01"))
)
kable(cohort, caption = "Example Cohort Table")
```

## Data Frame Structure

```{r}
#| echo: true
# Check structure
str(cohort)

# Access columns
cohort$subjectId
```

## Functions

```{r}
#| echo: true
# Functions are first-class objects in R
# Example: built-in functions
myMean <- mean
mySum <- sum

# They work the same as originals
testVector <- c(1, 2, 3, 4, 5)
print(paste("mean():", mean(testVector)))
print(paste("myMean():", myMean(testVector)))
```

## Lists: Complex Containers

```{r}
#| echo: true
# OHDSI packages often return lists
results <- list(
  cohortCounts = data.frame(
    cohortId = c(1, 2),
    cohortName = c("diabetes", "hypertension"),
    cohortCount = c(1520, 3842)
  ),
  executionTime = "2024-01-01",
  cdmSource = "Synpuf"
)

# Access with $ 
kable(results$cohortCounts, caption = "Cohort Counts from Results List")
```

## Accessing List Elements

```{r}
#| echo: true
# Multiple ways to access list elements
print(results$executionTime)
print(results[["cdmSource"]])
print(names(results))  # See what's in the list
```

# Working with Packages

## Installing OHDSI Packages

```r
# Install from CRAN
install.packages("DatabaseConnector")

# Install from GitHub (most OHDSI packages)
remotes::install_github("OHDSI/CohortGenerator")

# Load packages
library(DatabaseConnector)
library(CohortGenerator)
```

## Package Namespacing: A Critical Concept

- Multiple packages can have functions with the same name
- This causes **conflicts** that can break your code
- OHDSI packages often conflict with tidyverse packages

```{r}
#| echo: true
# Example of conflicting functions
# Both dplyr and stats have a filter() function

# After loading dplyr, filter() refers to dplyr's version
# This can break code expecting stats::filter()
```

## The Double Colon (::) Operator

```{r}
#| echo: true
# Use :: to explicitly specify which package
# PackageName::functionName()

# Examples with real functions
base::mean(c(1,2,3,4,5))
base::sum(c(1,2,3,4,5))

# This is especially important with OHDSI packages
# DatabaseConnector::connect()
# CohortGenerator::generateCohortSet()
```

## Common Namespace Conflicts in OHDSI

```{r}
#| echo: true
library(knitr)
conflicts_table <- data.frame(
  Function = c("select()", "filter()", "summarize()", "sql()"),
  `Package 1` = c("dplyr", "dplyr", "dplyr", "SqlRender"),
  `Package 2` = c("MASS", "stats", "Hmisc", "DBI"),
  Resolution = c("dplyr::select()", "dplyr::filter()", "dplyr::summarize()", "SqlRender::sql()")
)
kable(conflicts_table, caption = "Common Function Conflicts", col.names = c("Function", "Package 1", "Package 2", "Best Practice"))
```

## Checking for Conflicts

```{r}
#| echo: true
#| eval: false
# See all conflicts in your session
conflicts()

# See which package a function comes from
environment(mean)  # Shows it's from base/stats

# Check what's loaded
search()  # Shows all attached packages
```

## Best Practices for Namespacing

::: {.columns}
::: {.column width="50%"}
### Always Use :: For
- Critical functions
- Functions from multiple packages
- OHDSI package functions
- When sharing code
:::

::: {.column width="50%"}
### Benefits
- Code is self-documenting
- Prevents mysterious errors
- Works regardless of load order
- Makes dependencies clear
:::
:::

## Example: OHDSI with Explicit Namespacing

```r
# Clear and unambiguous
connection <- DatabaseConnector::connect(connectionDetails)

cohorts <- CohortGenerator::generateCohortSet(
  connection = connection,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTable = cohortTable
)

results <- CohortMethod::runCohortMethod(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  exposureDatabaseSchema = exposureDatabaseSchema
)

DatabaseConnector::disconnect(connection)
```

## Managing Dependencies

- OHDSI packages have many dependencies
- Use `renv` for reproducibility
- Check package versions when debugging
- Especially useful for working with SAFE Desktop

```{r}
#| echo: true
#| eval: false
# Check loaded packages
sessionInfo()

# See package version
packageVersion("DatabaseConnector")

# See dependencies of a package
tools::package_dependencies("CohortGenerator")
```

## Common OHDSI Package Patterns

```r
# 1. Create connection
connection <- connect(connectionDetails)

# 2. Execute analysis
results <- runAnalysis(
  connection = connection,
  cdmDatabaseSchema = "cdm",
  cohortDatabaseSchema = "results"
)

# 3. Clean up
disconnect(connection)
```

# Debugging Tips

## Reading Error Messages

```r
Error in connectUsingDriver(dbms = dbms, ...) : 
  Could not connect to database:
  FATAL: password authentication failed for user "ohdsi"
```

**Key parts:**

- Function where error occurred
- Specific error message
- Often includes hints for fixing

## Debugging Strategies

1. **Read the full error message**
2. **Check your connection:** `connection <- connect(cd)`
3. **Verify schemas exist:** Check cdmDatabaseSchema
4. **Look at intermediate results:** Use `str()` and `head()`
5. **Check the forums:** forums.ohdsi.org

## Useful Debugging Functions

```{r}
#| echo: true
# Create sample data for demonstration
debugResults <- list(
  data = data.frame(
    id = 1:5,
    value = c(10, 20, 30, 40, 50)
  ),
  status = "complete"
)

# Examine object structure
str(debugResults)
```

## More Debugging Functions

```{r}
#| echo: true
# Check object type
class(debugResults)
class(debugResults$data)

# List object contents
names(debugResults)

# Check if object exists
exists("debugResults")
exists("nonExistentObject")
```

## Using head() and tail()

```{r}
#| echo: true
# Create larger dataset
largeCohort <- data.frame(
  cohortId = rep(1:3, each = 10),
  subjectId = 101:130,
  eventDate = seq(as.Date("2024-01-01"), by = "day", length.out = 30)
)

# View first and last rows
kable(head(largeCohort, 3), caption = "First 3 rows")
kable(tail(largeCohort, 3), caption = "Last 3 rows")

```

# File I/O

## Reading and Writing CSV

```{r}
#| echo: true
# Create example data to write
cohortData <- data.frame(
  cohortId = c(1, 2, 3),
  cohortName = c("Type 2 Diabetes", "Hypertension", "Depression"),
  conceptId = c(201826, 316866, 440383)
)

# Write to temporary file for demo
tempFile <- tempfile(fileext = ".csv")
write.csv(cohortData, file = tempFile, row.names = FALSE)

# Read it back
readData <- read.csv(tempFile)
kable(readData, caption = "Data read from CSV")
```

## Working with JSON

```{r}
#| echo: true
library(jsonlite)

# Create example cohort definition structure
cohortDef <- list(
  id = 1,
  name = "Type 2 Diabetes",
  expression = list(
    conceptId = 201826,
    includeDescendants = TRUE
  )
)

# Convert to JSON
jsonString <- toJSON(cohortDef, pretty = TRUE, auto_unbox = TRUE)
cat(jsonString)
```

# Best Practices

## Code Style for OHDSI

- Use **camelCase** for variables (OHDSI standard)
- Clear, descriptive names
- Comment complex logic
- Format code consistently

```r
# Good
targetCohortId <- 1
comparatorCohortId <- 2

# Less clear
t <- 1
c <- 2
```

## Project Organization

```
my-study/
├── R/              # R scripts
├── inst/
│   ├── cohorts/    # Cohort JSONs
│   └── sql/        # SQL queries
├── results/        # Output files
└── extras/         # Additional scripts
```

## Version Control

- Always use Git for your projects
- Commit frequently with clear messages
- Never commit passwords or data
- Use `.gitignore` for sensitive files

```bash
# Basic git workflow
git add .
git commit -m "Add cohort diagnostics"
git push
```

# Summary

## Key Takeaways

- **R is essential** for OHDSI work
- **Understand the basics** before diving into packages
- **AI helps but verify** - always test generated code
- **Practice debugging** - errors are learning opportunities
- **Follow conventions** - makes collaboration easier


## Essential Skills Checklist

```{r}
#| echo: false
library(knitr)
checklist <- data.frame(
  Skill = c(
    "Create and use R projects",
    "Work with vectors and data frames",
    "Install and load packages",
    "Use package namespacing (::)",
    "Read error messages",
    "Debug with str(), head(), class()",
    "Read and write CSV/JSON files",
    "Use version control"
  ),
  Importance = c(
    "Sets working directory, organizes files",
    "Core data structures in R",
    "Access OHDSI tools",
    "Avoid conflicts and errors",
    "First step in troubleshooting",
    "Understand your data",
    "Import/export results",
    "Track changes, collaborate"
  )
)
kable(checklist, col.names = c("Essential Skill", "Why It Matters"))
```

## Resources

- [R for Data Science](https://r4ds.had.co.nz/)
- [OHDSI HADES](https://ohdsi.github.io/Hades/)
- [OHDSI Forums](https://forums.ohdsi.org/)
- [OHDSI Code Style](https://ohdsi.github.io/Hades/codeStyle.html)

## Questions?

Thank you!

Practice makes perfect - start small and build up!